Retrofitting Effect Handlers onto OCaml

### 2021-03-11 retro-concurrency.tex
Landscape mode supplement figures

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    %\vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.44]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
\vspace{-2mm}
\caption{Program stack on stock OCaml.}
\vspace{-2mm}
\end{figure*}

@@

\vspace{-2mm}

@@

\vspace{-1mm}
\subsubsection{Top-level reductions}

@@

\vspace{-1mm}
\subsubsection{C reductions}

@@

\vspace{-1mm}
\subsubsection{OCaml reductions}

@@

\vspace{-2mm}
\subsection{Heap-allocated fibers}
\label{sec:hafibers}

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
	with | v -> v | effect F kf -> ()
with | v -> v
		 | effect E ke -> (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
\vspace{-3mm}
\end{figure*}

@@

\vspace{-3mm}
\subsection{External calls and callbacks}

@@

\vspace{-3mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\vspace{-3mm}
\subsection{Stack unwinding}

@@

\vspace{-2mm}
\subsection{Garbage collection}

@@

The complete results are presented in the supplementary material, which we
summarise here. On average (geometric mean of the normalized values against
|stock| as the baseline), the multicore variants were \textbf{less than 1\%
slower} than |stock|. The outliers (on either ends) were due to the difference
in the allocator and the GC between stock and Multicore OCaml.

The biggest impact was the increase in the OCaml text section size (OTSS) due
to the stack overflow checks. We define OTSS as the sum of the sizes of all the
OCaml text sections in the compiled binary file ignoring the data sections, the
debug symbols, the text sections associated with OCaml runtime and other
statically linked C libraries. Compared to |stock|, OTSS is 19\% more for |MC|
and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
16-word red zone is effective at reducing OTSS compared to having no red zone,
whereas the 32-word red zone does not noticeably reduce OTSS further. Further
work is required to bring OTSS closer to |stock|.

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
\vspace{-7mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
\vspace{-5mm}
}
\label{tab:micro_noeffect}
\end{table}

@@

\vspace{-3mm}
\subsection{No perform benchmarks}

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 12.25 & 12.05 & 11.44 & 6.74 & 8.9 \\
	\texttt{monad} 	& 348.69 & 69.77 & 39.24 & 33.29 & 42.79 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	\vspace{-5mm}
\end{table}

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
		\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
		\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\vspace{-3mm}
	\caption{Web server performance.}
	\vspace{-5mm}
\end{figure}

@@

\vspace{-2mm}
\section{Related Work}
\label{sec:related}

@@

\vspace{-3mm}
\section{Conclusions}
\label{sec:conc}

@@

@@

\subsection*{No effect macro benchmarks}

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
	\label{res:macro_size}
\end{figure*}

Figures~\ref{res:macro_time} and \ref{res:macro_size} present the impact on
macro benchmarks that do not use effects using the Multicore OCaml compiler
which supports effect handlers. The aim is to quantify the impact of the
addition of effect handlers to the language on code that does not use effect
handlers. The |MC| is Multicore OCaml with the default red zone size (16
words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
overflow check. |MC+RedZone32| uses a red zone size of 32 words.

>>>>

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    %\vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.44]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
%\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
%\vspace{-2mm}
\caption{Program stack on stock OCaml.}
%\vspace{-2mm}
\end{figure*}

@@

%\vspace{-2mm}

@@

%\vspace{-1mm}
\subsubsection{Top-level reductions}

@@

%\vspace{-1mm}
\subsubsection{C reductions}

@@

%\vspace{-1mm}
\subsubsection{OCaml reductions}

@@

%\vspace{-2mm}
\subsection{Heap-allocated fibers}
\label{sec:hafibers}

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    %\vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
		\vspace{1mm}
    \begin{lstlisting}
effect E : unit
effect F : unit;;

match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
	with | v -> v | effect F kf -> ()
with | v -> v
		 | effect E ke -> (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
		\vspace{2mm}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
%\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
%\vspace{-3mm}
\end{figure*}

@@

%\vspace{-3mm}
\subsection{External calls and callbacks}

@@

%\vspace{-3mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

%\vspace{-3mm}
\subsection{Stack unwinding}

@@

%\vspace{-2mm}
\subsection{Garbage collection}

@@

\begin{figure*}
	\includegraphics[scale=0.35]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in kilobytes given in parenthesis.}
	\label{res:macro_size}
\end{figure*}

Figure~\ref{res:macro_time} presents the running time of the different
multicore variants normalized against the sequential baseline |stock|. On
average (geometric mean of the normalized values against |stock| as the
baseline), the multicore variants were \textbf{less than 1\% slower} than
|stock|. The outliers (on either ends) were due to the difference in the
allocator and the GC between stock and Multicore OCaml.

The biggest impact was the increase in the OCaml text section size (OTSS) due
to the stack overflow checks. We define OTSS as the sum of the sizes of all the
OCaml text sections in the compiled binary file ignoring the data sections, the
debug symbols, the text sections associated with OCaml runtime and other
statically linked C libraries. Figure~\ref{res:macro_size} presents the OTSS of
the multicore variants normalized against the sequential baseline |stock|.
Compared to |stock|, OTSS is 19\% more for |MC| and |MC+RedZone32|, and 30\%
more for |MC+RedZone0|. The result shows that our 16-word red zone is effective
at reducing OTSS compared to having no red zone, whereas the 32-word red zone
does not noticeably reduce OTSS further. Further work is required to bring OTSS
closer to |stock|.

\begin{table}
\caption{Micro benchmarks without effects. Each entry is the percentage
	difference for Multicore OCaml over stock OCaml.}
%\vspace{-7mm}
{
\begin{tabular}{r c c c c c c c c c c}
	& \rot{exnval} & \rot{exnraise} & \rot{extcall} & \rot{callback} & \rot{ack}
	& \rot{fib} & \rot{motzkin} & \rot{sudan} & \rot{tak} \\ \hline
	\textbf{Time} & +0.0 & -1.9 & +17 & +65  & +5.3
								& +2.2 & +10 & +0.0 & +4.2 \\
	\textbf{Instr} & +0.0 & +0.0 & +10 & +72 & +16
								 & +24 & +16 & +14 & +17 \\ \hline
\end{tabular}
%\vspace{-5mm}
}
\label{tab:micro_noeffect}
\end{table}

@@

%\vspace{-3mm}
\subsection{No perform benchmarks}

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
%\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 12.25 & 12.05 & 11.44 & 6.74 & 8.9 \\
	\texttt{monad} 	& 348.69 & 69.77 & 39.24 & 33.29 & 42.79 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	%\vspace{-5mm}
\end{table}

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
		%\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
		%\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	%\vspace{-3mm}
	\caption{Web server performance.}
	%\vspace{-5mm}
\end{figure}

@@

%\vspace{-2mm}
\section{Related Work}
\label{sec:related}

@@

%\vspace{-3mm}
\section{Conclusions}
\label{sec:conc}

@@

%\clearpage

@@
